Model Finding for Recursive Functions in SMT

### 2016-01-26 ijcar2016/body.tex
more on Nun

\hyphenation{counter-example counter-examples Isa-belle}

@@

Nunchaku is a new model finder, intended
to be used for finding counter-examples in several proof assistants. The
first version, 0.1, was released in January 2016 with support for (co)datatypes,
(co)recursive functions and (co)inductive predicates.
Support for higher-order
functions is planned in the next release.
We intend for Nunchaku to be supported at least in Isabelle (where basic
support for it already exists), Coq, and the TLA\textsuperscript{+} proof
system.

Nunchaku is the spiritual successor to Nitpick~\cite{blanchette-nipkow-2010},
but is developed as a standalone OCaml program (with its own input language) so
as to be usable independently from Isabelle.
Whereas Nitpick relies on propositional checking with Kodkod,
generating a succession of problems where cardinalities of finite types
grow at each step, Nunchaku
translates its input to one first-order logic formula that targets
the finite-model finding fragment of CVC4, including the (co)datatypes support.
Using CVC4 will also allow Nunchaku to provide efficient arithmetic and makes
equality reasoning straightforward.

The following problem is expressed in Nunchaku's input syntax; it starts with
a datatype declaration, then two mutually recursive coinductive predicates
are defined, and finally a distinguished axiom called the goal
(``find a non-zero natural number $m$ that is even'').
\begin{quote}
\begin{verbatim}
data nat := zero | Suc nat.

copred even : nat -> prop :=
  even zero;
  forall (n : nat). odd n => even (Suc n)
and odd : nat -> prop :=
  forall (n : nat). even n => odd (Suc n).

goal exists m. (even m && ~ (m = zero)).
\end{verbatim}
\end{quote}

Nunchaku parses and types the input problem before applying a sequence
of translations, each reducing the distance to the target fragment. Here,
for instance, the coinductive predicates
\verb!even! and \verb!odd! are \emph{polarized} (specialized into
\verb!even+!, used only in positive positions, and \verb!even-! used in negative
positions),
$m$ is Skolemized into a fresh constants,
then then translated into admissible functions
before another pass uses the encoding from Section~\ref{sec:encoding}.
If a model is found, it is translated back to the input language.
Conceptually, the sequence of transformation is a \emph{two ways pipeline}
built by composing pairs \textsf{(encode,decode)}.
For each such pair, \textsf{encode}
transforms a problem over a signature $\Sigma$ in a logic
$\mathcal{L}$ to a problem over a signature $\Sigma'$ in a slightly different
logic $\mathcal{L}'$, and \textsf{decode} translates a model of $\mathcal{L'}$
over $\Sigma'$ into a model of $\mathcal{L}$ over $\Sigma$. The pipeline
currently implemented in Nunchaku is described below.

\begin{description}
%{{{
  \item[Type inference] infers types and checks definitions;
  \item[Type skolemization]
    replaces $\exists \alpha. p[\alpha]$
      with $p[\tau]$ where $\tau$ is a fresh type;
  \item[Monomorphization]
    specializes polymorphic functions on their type arguments;
  \item[Elimination of Equations]
    translates multiple-equations definitions of functions into
      a single nested pattern matching;
  \item[Polarization]
    specializes some predicates into a version used in positive position
      and one used in negative position;
  \item[Unrolling]
    adds a decreasing argument to possibly ill-founded predicates;
  \item[Skolemization]
    introduces Skolem symbols for term variables;
  \item[Elimination of Predicates]
    changes a multi-clauses (co)inductive predicate definition
      into a single recursive function;
  \item[Recursion Elimination]
    performs the encoding from Section~\ref{sec:encoding};
  \item[Elimination of Match]
    transforms pattern-matching expressions into tests and selectors;
  \item[CVC4 Invocation] calls CVC4 to obtain a model.
%}}}
\end{description}

TODO: similar to institutions \cite{goguen-meseguer} (?)

\subsection{Decoding of Recursive Functions}

Nunchaku is designed for finding counter-models to formulas a user thinks
are theorems. Decoding models from CVC4 to hide any trace of the successive
encodings is therefore vital for readability. Each step
of the pipeline only has to concern itself with removing its changes to
the signature and logic; our encoding of recursive functions is no exceptions.

TODO explain how:
- domains of proj. functions are computed
- domains of original function is deduced
- decision tree for original function is built
- note how junk is removed in step 3 (but undefined "..." as last case)

\subsection{Support for Higher-Order Functions}

TODO:
- currying with explicit application symbol --> function arguments are now
  just constants
- apply encoding on application symbol
- gotcha: one function can be used with several arities.

- optim: specialization (similar to monomorphization)

>>>>

\hyphenation{counter-example counter-examples Isa-belle data-type data-types
  co-data-type co-data-types}

@@

Nunchaku is a new higher-order model finder designed to be integrated with
several proof assistants. The first version, 0.1, was released in January 2016
with support for (co)datatypes, (co)recursive functions and (co)inductive
predicates. Support for higher-order functions is planned for the next
release. We have developed a preliminary Isabelle frontend and are planning
further frontends for Coq, the TLA\textsuperscript{+} Proof System, and
possibly other systems.

Nunchaku is the spiritual successor to Nitpick for
Isabelle/HOL~\cite{blanchette-nipkow-2010}, but is developed as a standalone
OCaml program, with its own input language. Whereas Nitpick generates a
succession of problems where cardinalities of finite types grow at each step,
Nunchaku translates its input to one first-order logic program that targets
the finite model finding fragment of CVC4, including interpreted
(co)datatypes \cite{reynolds-blanchette-2015-codata}.
Using CVC4 also allows Nunchaku to provide efficient arithmetic
reasoning and to detect unsatisfiability in addition to satisfiability.
% and makes
% equality reasoning straightforward.

The input syntax was inspired by that other systems based on higher-order
logic (e.g., Isabelle/HOL) and by functional programming languages (e.g.,
OCaml). The following simple problem gives a taste of the syntax:
%
\begin{quote}
\verb$data nat := Zero | Suc nat.$ \\[2\jot]
\verb$pred even : nat -> prop :=$ \\
\verb$  even Zero;$ \\
\verb$  forall n. odd n => even (Suc n)$ \\
\verb$and odd : nat -> prop :=$ \\
\verb$  forall n. even n => odd (Suc n).$ \\[2\jot]
\verb$val m : nat.$ \\
\verb$goal even m && ~ (m = Zero).$
\end{quote}
%
The problem defines a datatype (\texttt{nat}) and two mutually recursive
inductive predicates (\texttt{even} and \texttt{odd}), it declares a constant
\texttt{m}, and specifies a goal to satisfy (``$m$ is even and nonzero'').
Nunchaku quickly finds the following partial model:
%
\begin{quote}
\verb$val m := Suc (Suc Zero).$ \\
\verb$val odd := fun x. if x = Suc Zero then true else ?__.$ \\
\verb$val even := fun x. if x = Suc (Suc Zero) || x = Zero then true$ \\
\verb$                   else ?__.$
\end{quote}
%
The partial model gives sufficient information to the user to evaluate the
goal: ``2 is even if 1 is odd, 1 is odd if 0 is even, and 0 is even.''

Nunchaku parses and types the input problem before applying a sequence
of translations, each reducing the distance to the target fragment.
In our example, the predicates
\verb!even! and \verb!odd! are \emph{polarized} (specialized into a pair of
predicates such that one is used in positive positions and the other in
negative positions),
$m$ is skolemized into a fresh constants,
then translated into admissible recursive functions,
before another pass applies the encoding described in this paper.
If a model is found, it is translated back to the input language,
with \verb$?__$ placeholders as necessary.

Conceptually, the sequence of transformation is a \relax{two-way pipeline}
built by composing pairs $(\textsf{encode}, \textsf{decode})$ of
transformations. For each such pair, \textsf{encode} transforms a problem over
a signature $\Sigma$ in a logic $\mathcal{L}$ to a problem over a signature
$\Sigma'$ in a logic $\mathcal{L}'$, and \textsf{decode} translates a model in
$\mathcal{L'}$ over $\Sigma'$ into a model in $\mathcal{L}$ over $\Sigma$,
in the spirit of institution theory \cite{goguen-burstall-1992}. The
pipeline currently implemented in Nunchaku consists of the following phases:

\begin{description}
%{{{
  \item[Type inference] infers types and checks definitions;
  \item[Type skolemization]
    replaces $\exists \alpha.\; \varphi[\alpha]$
      with $\varphi[\tau]$, where $\tau$ is a fresh type;
  \item[Monomorphization]
    specializes polymorphic definitions on their type arguments and removes
    unused definitions;
  \item[Elimination of equations]
    translates multiple-equation definitions of recursive functions into
      a single nested pattern matching;
  \item[Polarization]
    specializes predicates into a version used in positive positions
      and a version used in negative positions;
  \item[Unrolling]
    adds a decreasing argument to possibly ill-founded predicates to make
    them well founded;
  \item[Skolemization]
    introduces Skolem symbols for term variables;
  \item[Elimination of Predicates]
    changes a multiple-clauses (co)inductive predicate definition
      into a single recursive function;
  \item[Recursion elimination]
    performs the encoding from Section~\ref{sec:encoding};
  \item[Elimination of pattern matching]
    rewrites pattern-matching expressions using datatype
    discriminators and selectors;
  \item[CVC4 invocation] runs CVC4 to obtain a model.
%}}}
\end{description}

% \subsection{Decoding of Recursive Functions}
% 
% Nunchaku is designed for finding counter-models to formulas a user thinks
% are theorems. Decoding models from CVC4 to hide any trace of the successive
% encodings is therefore vital for readability. Each step
% of the pipeline only has to concern itself with removing its changes to
% the signature and logic; our encoding of recursive functions is no exceptions.
% 
% TODO explain how:
% - domains of proj. functions are computed
% - domains of original function is deduced
% - decision tree for original function is built
% - note how junk is removed in step 3 (but undefined "..." as last case)
% 
% \subsection{Support for Higher-Order Functions}
% 
% TODO:
% - currying with explicit application symbol --> function arguments are now
%   just constants
% - apply encoding on application symbol
% - gotcha: one function can be used with several arities.
% 
% - optim: specialization (similar to monomorphization)
